---
title: "TRRident: Identifying T Cell Entry Regions in Spatial Transcriptomics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TRRident: Identifying T Cell Entry Regions in Spatial Transcriptomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction
TRRident is a computational framework for identifying T cell entry regions (TRR) in spatial transcriptomics data. The method uses Ising models with impurities to detect regions where T cells enter the tumor core.

## Installation

```{r installation}
# Install from GitHub
# devtools::install_github("yourusername/TRRident")

# Load the package
library(TRRident)
```

## Data Requirements

TRRident works with 10x Visium spatial transcriptomics data. You need:

1. **HDF5 filtered feature-barcode matrix** (e.g., `filtered_feature_bc_matrix.h5`)
2. **Spatial imaging data folder** (must be named `spatial/`)
3. The `tissue_positions.csv` file from the spatial folder

## Tutorial: Melanoma Analysis

### Step 1: Load Data

First, define your marker genes:

```{r markers}
# T cell markers
tcell_markers <- c('CD3D', 'CD3E', 'CD3G', 'CD4', 'CD8A', 'CD8B2')

# Exhaustion markers
ex_markers <- c('PDCD1', 'LAG3', 'HAVCR2', 'TIGIT', 'ENTPD1', 'CTLA4', 'TOX')
```

Load and prepare the data:

```{r load_data}
dat1 <- data_preparation_TRR(
  folder = 'data/visium_melanoma1',
  filenm = "CytAssist_FFPE_Human_Skin_Melanoma_filtered_feature_bc_matrix.h5",
  res = 0.2,
  loc_file = 'data/visium_melanoma1/spatial/tissue_positions.csv',
  ex_markers = ex_markers,
  tcell_markers = tcell_markers
)

# Extract prepared data
sp.counts.r <- dat1[[1]]
sp.counts.norm <- dat1[[2]]
sp.counts.tpm <- dat1[[3]]
vgg <- dat1[[4]]
loc.raw <- dat1[[5]]
meta_dat <- dat1[[6]]
r.unit <- dat1[[7]]
```

### Step 2: Identify Tumor Clusters

```{r tumor_clusters}
# Load cancer markers (you need to provide these)
# cancer.markers$melanoma should contain melanoma-specific genes

tmp.mean <- plot_seurat_clusters(
  loc = loc.raw,
  meta_dat = meta_dat,
  sp.counts.norm = sp.counts.norm,
  sp.counts.r = sp.counts.r,
  cancer.markers = cancer.markers$melanoma
)

# Define tumor clusters based on high cancer marker expression
tumor.clus <- c(0, 4, 5, 6)  # Adjust based on your data
tumor.loc <- loc.raw[rownames(meta_dat)[which(meta_dat$seurat_clusters %in% tumor.clus)], ]
```

### Step 3: Estimate J1 for the Ising Model

J1 is the coupling parameter that controls how strongly neighboring spots influence each other:

```{r estimate_j1}
j1.dat <- estimate_J1_dat(genes = 'B2M', sp.dd = sp.counts.r, adj.umi = FALSE)

sum_obs <- j1.dat[[2]]
j1.dat <- j1.dat[[1]]

loc.dat <- j1.dat[[2]]
obs_ss <- j1.dat[[3]]
nb.r <- j1.dat[[4]]
nb.info <- j1.dat[[5]]

# Find optimal J1 using variational inference
first_dr <- find_optimal_mu_Q(
  mu_Q_candidates = seq(0.16, 0.44, by = 0.01),
  init_x = sample(c(-1, 1), size = length(obs_ss), prob = c(0.5, 0.5), replace = TRUE),
  nb.list.cpp = nb.info,
  mu = 0.3, sigma2 = 0.0049, sigma2_Q = 0.0001,
  K = 10, s_obs = sum_obs, n_flip = 60000, M = 10000
)

J1 <- as.numeric(names(first_dr)[which.min(first_dr)])
```

### Step 4: Decompose 2D Data into 1D Lines

```{r get_lines}
all.lines <- get_lines(loc.raw, grid.type = 'hex', max_gap = 7, aa = 0.2)
plot_lines(loc.raw, all.lines, grid.type = 'hex')
```

### Step 5: Get Margin Spots

```{r margin_spots}
margin.spots <- c()
for (i in 1:length(all.lines)) {
  tmp.lines <- all.lines[[i]]
  margin.spots <- c(margin.spots, unlist(lapply(tmp.lines, function(x) 
    rownames(x)[c(1, nrow(x))])))
}
margin.spots <- unique(margin.spots)

plot(loc.raw)
points(loc.raw[margin.spots, ], col = 'red')
```

### Step 6: Detect Change Points

```{r change_points}
NN <- nrow(sp.counts.r)
YY <- rowSums(sp.counts.r[, ex_markers])
XX <- cbind(rep(1, NN), rep(NA, NN), rowSums(sp.counts.r[, tcell_markers]))
rownames(XX) <- names(YY)

all.output.tcell <- get_cps_and_save(
  all.lines = all.lines,
  YY = YY,
  XX = XX,
  K = 0,
  alpha_vec = seq(0.1, 1, length = 10),
  method_cp = 'PG',
  dat_name = 'data/melanoma1_tcelladj',
  grid.type = 'hex'
)
```

### Step 7: Calculate Slopes

```{r slopes}
slope.output <- calculate_slopes(
  all_segs = all.segs.list[[10]],
  all.lines = all.lines,
  y_markers = ex_markers,
  sp.counts.r = sp.counts.r,
  method_beta = 'PG',
  alpha = alpha_vec[10],
  dat_name = 'data/melanoma1'
)
```

### Step 8: Estimate J2

J2 controls the impurity interactions at change points:

```{r estimate_j2}
lambda.list <- seq(1, 10, length = 5)
j2.dat <- plot_J2(J1 = J1, lambda.list = lambda.list)

mean.list <- j2.dat[[1]]
n.add <- j2.dat[[2]]
new.nb.list <- j2.dat[[3]]

# Plot to find transition point
plot(mean.list, type = 'l', ylim = c(0, 20), xaxt = "n", 
     xlab = "J2", ylab = '% spots with Pr(TRR)>0.7')
axis(1, at = seq_along(mean.list), labels = round(lambda.list, 1))

# Define J2 at the transition point
J2 <- 5  # Adjust based on the plot
```

### Step 9: Gibbs Sampling

```{r gibbs}
ss.mat <- gibbs_sampling_new(
  ss = ss,
  spot_names = names(obs_ss),
  n_add = n.add * J2,
  nb_list = nb.r,
  new_nb_list = new.nb.list,
  J = J1,
  loc = loc.raw,
  iterations = 5000
)

ss.mat[ss.mat == -1] <- 0
final.probs <- calculate_final_prob(ss.mat, loc = loc.dat, burn_in = 2000, J = J1)
```

### Step 10: Get Final TRR Regions

```{r get_trr}
final.regions <- get_TRR(final.probs)
```

## Downstream Analysis

After identifying TRR regions, you can perform differential expression analysis:

```{r downstream}
# Define TRR spots and control spots
trr_spots <- unlist(final.regions)
control_spots <- rownames(tumor.loc)[!rownames(tumor.loc) %in% trr_spots]

# Differential expression with T cell marker adjustment
de_results <- gene_comparison_tmaker_adj_all(
  genes = vgg,
  ids0 = control_spots,
  ids1 = trr_spots,
  umi.adj = TRUE
)

# Filter significant genes
sig_genes <- de_results[de_results$adj_p_value < 0.05 & abs(de_results$logFC) > 1, ]
```

## Session Info

```{r session}
sessionInfo()
```
